Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-01-22 src/article.tex
Updates from Overleaf

The design of the augmented replica keyboard for the Tagliavini Collection was guided by three principal constraints stipulated by the museum:
\begin{itemize}
\item \emph{No visible electronics}. To preserve the visual integrity of the exhibit, all electronic components had to remain concealed except for a pair of headphones and a small display for audio parameter adjustments.
\item \emph{Robustness and reliability}. The system needed to accommodate frequent use by museum visitors and allow for straightforward maintenance by staff without requiring specialised technical expertise.
\item \emph{Faithfulness}. The keyboard mechanism ensured fidelity to its authentic operation, keeping artefacts to a minimum. 
\end{itemize}
These constraints demanded a non-invasive, robust design that was respectful of the historical authenticity of the instruments. The requirement for faithfulness to the original mechanism immediately ruled out a mechanical design reliant on electromechanical actuators, as in previous works on piano haptics \cite{Timmermans2020,Gillespie1996}. While actuators can generate considerable force, they cannot replicate the wideband signals characteristic of the harpsichord's sharp and responsive haptics (A REFERENCE WOULD BE GOOD HERE, ANY IDEAS?), which differ significantly from those of a piano.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/3-key-side-labelled.png}
    \caption{3-Key Model Harpsichord Mechanism designed by Graziano Bandini at San Colombano, Bologna, 2023. Key components include A: Wrestplank, B: Key Lever, C: Key Cap, D: Soundboard, E: Jack.}
    \label{fig:3key}
\end{figure}

@@

% \begin{figure*}
%     \centering
%     \includegraphics[width=\linewidth]{src/images/bl}
%     \caption{3-Key Model Harpsichord Mechanism designed by Graziano Bandini at San Colombano, Bologna, 2023. Key components include A: Wrestplank, B: Key Lever, C: Key Cap, D: Soundboard, E: Jack.}
%     \label{fig:3key}
% \end{figure*}

@@

Finer calibration carried out with the guidance of \anon{Catalina Vincens} as an expert harpsichord performer.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg}
    \caption{Cross-section illustrating sensor placement relative to the jack mechanism.}
    \label{fig:cross-section}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/baffles.png}
    \caption{STL Model of baffles designed to prevent cross-talk between adjacent sensors.}
    \label{fig:baffles}
\end{figure}

@@

Paper has presented the design for an interactive exhibition at a musical instrument museum. Design attempts to address the problem.
Optical sensors read by a microcontroller have been shown to be a solution to tracking vertical displacement of harpsichord jacks and therefore a viable system to augmenting harpsichords. The interface's capabilities of the jack rails to disengage and to convert string plucks to MIDI data means this can be used as a probe to explore the how the ``double-pluck'' can be used as vector for expression during performance. The placement of the instrument as a museum exhibit means it also has potential to be used in a long form study.

@@

A limitation in the design is the variability in positioning of tag,
especially when they need to be reapplied. A jig was designed for
applying tags, but could not be printed in time for installation. Even
with a jig, some tags had to be readjusted by hand regardless. Long term
this has implications on how to maintain the interface.

Taken a different approach to what surface is used. Preliminary test
were taken using Laser Cutting and CNC to fabricate Jack bodies. Laser
cutting results in a loss of material that has too much variability. The
edge of the jack body will be charred. The material also needs to be
reliable. on the advice of two luthiers, \anon{Roberto Livi} and \anon{Jonathan Santa
Maria Bouquet} at \anon{Edinburgh}, Plywood and MDF have problems with
variations in humidity, tear-out and delamination.

Some tests were carried out using a notched jack body that was 3D
printed. The notch means that as the jack raises, the forward material
acts like a shutter. No reasons that this could not be CNC though a
variety of materials would be needed. Traditional material used is xx
though need to test of the material is naturally reflective enough or if
a coating would need to be applied.

Outputs of the NEMUS project will also include numerically simulated
non-linear plucked string models with which the device can interact. A
bespoke sound synthesis engine to leverage data from both jack rows
which will also be open-sourced. Design for instrument to be compatible
with standard MIDI instruments as well software designed for the full
data output.

A second interface designed more closely to the Trasuntino style
instrument commissioned by the \anon{NEMUS} project with \anon{Roberto Livi}. Using
lessons learned for the first iteration the second can make better
accommodations for internal electronics. Interface to be used for a
performer study and used for performance with a numerically simulated
sound synthesis engine.

Testing current sensors in conjunction with force sensitive resistors to
research if a impulse response for the pluck can be extrapolated.

Future novel designs for an interface that would utilise velocity and
aftertouch messages in software. This would be applicable for numerical
simulations of strings. A bespoke piece of software would also allow
for direct mapping of jack rows to registers. There is enough data to
implement functionality for velocity and aftertouch MIDI messages and
simply needs to be activated in the firmware. (Hyperinstruments
\cite{nime2024_20})

seat vibrations \cite{MusicalHaptics2018_07}

\begin{itemize}
\item
  ``there is a general connection between vibrations and the perceived
  quality of music reproduction''
\item
  ``influence of vibrations on loudness perception at low frequencies''
  Evaluation
\end{itemize}

\subsection{improvements}\label{improvements}

\begin{itemize}
\item
  ESP-IDF

  \begin{itemize}
  \item
    latency half
  \item
    workload high
  \end{itemize}
\end{itemize}

A current limitation is the latency in triggering MIDI output. Currently
the latency is around 20 ms, but should ideally be closer to 5 ms.

An obvious approach would be merely changing the microcontroller. A test
was carried out using an Arduino Nano ESP32 which is a NORA-W106, a
module with a ESP32-S3. The benefit of this microcontroller is that it
would allow simply the board to be swapped and the rest of the PCB
designs would be unaffected.

Using the Espressif development framework ESP-IDF and following the same
logic as the original firmware source, the latency was reduced to around
10 ms. ESP-IDF allows access to functionality of the ESP32 that is not
exposed in the Arduino libraries. It is possible that the 10 ms latency
could be improved upon, but there would be a cost in the form of
development time. Consideration would need to be made on how to port
existing code or whether to completely re-write the firmware.

\begin{itemize}
\item
  avoid soldering cabling
\item
  power indicators on pcbs

  \begin{itemize}
  \item
    debug if problem is power related Add power small power LEDs to the
    each PCB. It was not common for a board to not be functioning and
    thye result to be that the power was not correctly connected.
    Especially when the power is connected via something such as an FFC
    cable where it is easy to plug-in the wrong way especially when
    fitting is to be carried out by someone unfamiliar with the system.
    Power LEDs would quickly confirm whether the problem was power
    related. They would have to be small enough and low enough current
    draw to avoid glare or being mistaken with other indicator LEDs To
    continue using THT components, the choice would have to be made
    carefully.
  \end{itemize}
\end{itemize}

>>>>

The design of the augmented replica keyboard for the Tagliavini Collection was guided by three principal constraints stipulated by the museum:
\begin{itemize}
\item \emph{No visible electronics}. To preserve the visual integrity of the exhibit, all electronic components had to remain concealed except for a pair of headphones and a small display for audio parameter adjustments.
\item \emph{Robustness and reliability}. The system needed to accommodate frequent use by museum visitors and allow for straightforward maintenance by staff without requiring specialised technical expertise.
\item \emph{Faithfulness}. The keyboard mechanism ensured fidelity to its authentic operation, keeping artefacts to a minimum. 
\end{itemize}
These constraints demanded a non-invasive, robust design that was respectful of the historical authenticity of the instruments. The requirement for faithfulness to the original mechanism immediately ruled out a mechanical design reliant on electromechanical actuators, as in previous works on piano haptics \cite{Timmermans2020,Gillespie1996}. While actuators can generate considerable force, they cannot replicate the wideband signals characteristic of the harpsichord's sharp and responsive haptics (A REFERENCE WOULD BE GOOD HERE, ANY IDEAS?), which differ significantly from a piano.

Here is a consolidated **Open-Source Design** subsection for your article:

\subsection{Open-Source Design and Cost-Effectiveness}\label{susec:open-source}

A central objective of this project was to ensure accessibility and reproducibility by committing to an open-source approach for all hardware, software, and data. The commitment to open-sourcing encompassed all aspects of the system, including hardware schematics, firmware, and calibration data. Cost-effectiveness was a central consideration. For example, the system was initially developed for a 3-key prototype and successfully scaled to 49 keys without significant increases in cost or complexity. Key components, such as QRE1113 sensors and CD4051BE multiplexers, were selected for their affordability and availability, while the modular PCB design ensured efficient replication and maintenance.

The Arduino Nano BLE was chosen as the core microcontroller for its compatibility with open-source tools and its ability to support both USB and BLE MIDI. Ferroelectric RAM (FRAM) was employed for data storage, offering an economical yet robust solution for preserving calibration settings across power cycles. Calibration workflows were optimised using the Arduino IDE’s serial plotter and open-source MIDI Monitor software, reducing reliance on proprietary tools and simplifying the process for users.

Scalability and ease of assembly were prioritised throughout the design process. The modular structure of the PCBs, combined with 3D-printed baffles and gradient stickers, ensured that the system could be scaled up without compromising performance or cost-efficiency. Installation and calibration times were reduced through iterative improvements, such as the integration of addressable LEDs for sensor identification during setup. These design choices align with the project’s open-source ethos, allowing assembly using standard tools commonly available in university maker spaces.

By making the hardware and software openly available, the project invites contributions and adaptations from the broader community. This transparency supports critical evaluation and encourages iterative improvements, promoting sustainability and innovation in interactive musical interface design. Moreover, the emphasis on affordability and accessibility ensures that the system can be adopted by a wide range of users, from cultural heritage institutions to academic researchers, reinforcing the project’s commitment to democratising access to advanced technological solutions.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg}
    \\
    \includegraphics[width=\linewidth]{src/images/3-key-side-labelled.png}
    \caption{3-Key Model Harpsichord Mechanism designed by Graziano Bandini at San Colombano, Bologna, 2023. Key components include A: Wrestplank, B: Key Lever, C: Key Cap, D: Soundboard, E: Jack.}
    \label{fig:3key}
\end{figure}

% \begin{figure}
%     \centering
%        \begin{subfigure}
%          \centering
%          \includegraphics[width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg}
%          \caption{Cross Section of 49-Key Model}
%          % \label{fig:y equals x}
%      \end{subfigure}\\
%     \begin{subfigure}
%          \centering
%          \includegraphics[width=\linewidth]{src/images/3-key-side-labelled.png}
%          \caption{3-Key Model: A: Wrestplank, B: Key Lever, C: Key Cap, D: Soundboard, E: Jack}
%          % \label{fig:y equals x}
%      \end{subfigure}        
%     \caption{Cross Sections of Models}
%     \label{fig:3key}
% \end{figure}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{src/images/system-block-diagram-long.png}
    \caption{Block diagram of PCB connection. Sensor and LED are routed through each sensor PCB}
    \label{fig:system-block-diagram}
\end{figure*}

@@

As an expert harpsichord performer, finer calibration was carried out with the guidance of \anon{Catalina Vincens}.

@@

% \begin{figure}
%     \centering
%     \includegraphics[width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg}
%     \caption{Cross-section illustrating sensor placement relative to the jack mechanism.}
%     \label{fig:cross-section}
% \end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/sensor-board-w-baffles.jpeg}
    \\
    \includegraphics[width=\linewidth]{src/images/baffles.png}
    \caption{Baffles designed to prevent cross-talk between adjacent sensors.}
    \label{fig:baffles}
\end{figure}

@@

Paper has presented the design for an interactive exhibition at a musical instrument museum. Design attempts to address the problem.
Optical sensors read by a microcontroller have been shown to be a solution to tracking vertical displacement of harpsichord jacks and therefore a viable system to augmenting harpsichords. The interface's capabilities of the jack rails to disengage and to convert string plucks to MIDI data means this can be used as a probe to explore how the ``double-pluck'' can be used as vector for expression during performance. The placement of the instrument as a museum exhibit means it also has potential to be used in a long form study.

@@

The stickers as a sensor surface have a few limitations that can be addressed. Firstly, the stickers need to be applied manually, which is laborious but also introduces variability. Secondly, the material for the stickers is rated at a 4 years lifespan before the adhesive will decay. The application is non-standard and it is likely there will be failure or drift in sensor readings over time, though to what extent is unknown. This presents a problem on how this approach can be maintained in the long term. Preliminary tests with 3D printed jack bodies incorporating the sensor surface were made. The jack body was notched and effectively acts a shutter against the optical sensor. More testing needs to be carried out before they are incorporated into the design. 3D printed jacks would solve two problems simultaneously in that they would be easily reproducible, but also remove labour demands on the luthier.

Their are also plans to develop the current setup by incorporating force sensitive resistors (FSR). 
FSRs would be used in conjunction with the current optical sensors to research if an impulse response for the pluck can be extrapolated.

% Taken a different approach to what surface is used. Preliminary test
% were taken using Laser Cutting and CNC to fabricate Jack bodies. Laser
% cutting results in a loss of material that has too much variability. The
% edge of the jack body will be charred. The material also needs to be
% reliable. on the advice of two luthiers, \anon{Roberto Livi} and \anon{Jonathan Santa
% Maria Bouquet} at \anon{Edinburgh}, Plywood and MDF have problems with
% variations in humidity, tear-out and delamination.

% Some tests were carried out using a notched jack body that was 3D
% printed. The notch means that as the jack raises, the forward material
% acts like a shutter. No reasons that this could not be CNC though a
% variety of materials would be needed. Traditional material used is xx
% though need to test of the material is naturally reflective enough or if
% a coating would need to be applied.

Space constraints meant that only one row of jacks could be sensed. A subsequent redesign in the dimensions of the PCB addressed this problem.
The redesign included flat flex cable (FFC) instead of header pins. This reduced the the size of the PCBs as well as the labour on hand soldering.
There was an additional added benefit in that the FFC has only a single orientation in which it conducts, meaning it cannot be incorrectly connected.

Currently the instrument communicates with a Spitfire sample library through the Kontakt software sampler. As such, the full data from both rows of jacks cannot be utilised. which is limited. A future output of the \anon{NEMUS} project will be a bespoke sound synthesis engine to leverage data from both jack rows
which will also be open-sourced. The intention is also to create a numerically simulated non-linear plucked string models with which the device can interact.

% The design for instrument to be compatible with standard MIDI instruments as well software designed for the full data output.

A second interface designed more closely to the Trasuntino style instrument has been commissioned by the \anon{NEMUS} project with \anon{Roberto Livi}. 
Using lessons learned for the first iteration the second can make better accommodations for internal electronics. Interface to be used for a performer study and used for performance with the aforementioned numerically simulated sound synthesis engine.

Future novel designs for an interface that would utilise available data, such as velocity and aftertouch, in software in the manner of a hyperinstrument.
Currently the system looks for the cross over a threshold for triggering MIDI messages, but jack displacement is tracked continuously. Data for velocity, both of the pluck and the jack can be extracted from the available data, simply requiring the addition of state machine to make use of it.
% This would be applicable for numerical simulations of strings.  
% A bespoke piece of software would also allow for direct mapping of jack rows to registers. 
% There is enough data to implement functionality for velocity and aftertouch MIDI messages and simply needs to be activated in the firmware. ()

Seat vibrations \cite{MusicalHaptics2018_07} can also be explored to determine if they would create a more immersive experience in the exhibition context.

% \begin{itemize}
% \item
%   ``there is a general connection between vibrations and the perceived
%   quality of music reproduction''
% \item
%   ``influence of vibrations on loudness perception at low frequencies''
%   Evaluation
% \end{itemize}

% \subsection{improvements}\label{improvements}

% \begin{itemize}
% \item
%   ESP-IDF

%   \begin{itemize}
%   \item
%     latency half
%   \item
%     workload high
%   \end{itemize}
% \end{itemize}

A current limitation is the latency in triggering MIDI output. Currently the latency from pressing a key to triggering a note is around 20 ms. 
The latency should ideally be closer to 10 ms and with no jitter so \cite{Jack2016}.

One approach would be to change the microcontroller. A test
was carried out using an Arduino Nano ESP32 which is a NORA-W106, a module containing an ESP32-S3. 
The benefit of this microcontroller is that it would allow the microcontroller to be swapped and the rest of the PCB
designs would be unaffected.

Another approach would be to use a different microcontroller entirely. 
Some research was carried during the course of the project using an ESP32 programmed with the Espressif development framework ESP-IDF. 
The same logic as the original firmware was programmed and the latency was reduced to below 10 ms. ESP-IDF allows access to functionality of the ESP32 including regular processing of an internal audio buffer that is not exposed in the Arduino IDE. 
It is possible that the 10 ms latency could be improved upon with appropriate optimisation, but there would be a cost in the form of development time. 
Consideration would need to be made on the time to port existing source code for the I2C FRAM chip or fall back on a simpler storage option.

% \begin{itemize}
% \item
%   avoid soldering cabling
% \item
%   power indicators on pcbs

%   \begin{itemize}
%   \item
%     debug if problem is power related Add power small power LEDs to the
%     each PCB. It was not common for a board to not be functioning and
%     thye result to be that the power was not correctly connected.
%     Especially when the power is connected via something such as an FFC
%     cable where it is easy to plug-in the wrong way especially when
%     fitting is to be carried out by someone unfamiliar with the system.
%     Power LEDs would quickly confirm whether the problem was power
%     related. They would have to be small enough and low enough current
%     draw to avoid glare or being mistaken with other indicator LEDs To
%     continue using THT components, the choice would have to be made
%     carefully.
%   \end{itemize}
% \end{itemize}
